---
title: "Visualization part 2"
author: "Shihui Peng"
date: "2023-10-12"
output: github_document
---
```{r, message=FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

# Get the data for plotting 
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

## labeling w `labs()`
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  )
```
* **`labs()`**: label, and can label lots of aesthetics. inside the `labs()`:
  * `x =` & `y =`: label the x and y axis.
  * `color =`: we have a color aeshetics mapped onto the name var, so the name var show up as the legend of the color bar. now, i want to label color as 'location'.
  * `title=` and `caption=` to define title and caption
  
## scales

we have basically 1 to 1 mapping of whatever on the aes to whatever var that are put in. but we might want to do some tramsformation in some association. we can start from adding diff scales to it to modify the default mapping bt a var and the aes
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  scale_x_continuous(
    breaks = c(-15, 0, 15),
    labels= c('-15C', '0C', '15C')
  ) +
  scale_y_continuous(
    position = 'right',
    trans = 'sqrt',
    limit = c(20, 30)
  )
```

* we can control over the location and specification of tick marks on the X or Y axis w **`scale_x_* and scale_y_*`** where * depends on the type of variable mapped to the x and y aesthetics (i.e. continuous vs discrete).
  * **`scale_x_continuous()`**: 
    * `breaks = c(-15, 0, 15)`: use this to control where things fall on the x-axis
    * `labels =` is to modify the labels for each point falls o the x-axis.
  * **`scale_y_continuous()`**:
    * `position = 'right'`: move the info of y-axis from left (default) to right.
    * `trans = 'sqrt'`: do a square root transformation. we can also do this w `mutate` before ggplot, and this can be embedded to further down below.
    * `limit = c(20, 30)`: limit the y-axis value wi 20~30. but `filter()` might be better, bc w limit=, x-axis here is not updated and is still the corresponding range with the previous y range.
    
## colors
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  viridis::scale_color_viridis(discrete = TRUE)
  
```
* we can use `scale_color_hue(h = c(150, 300))` to create our own color scheme, but it usually does not go well. So use **viridis** package is better.
* We used `discrete = TRUE` because the color aesthetic is mapped to a discrete variable. In other cases (for example, when color mapped to prcp) you can omit this argument to get a continuous color gradient. The `viridis::scale_fill_viridis()` function is appropriate for the fill aesthetic used in histograms, density plots, and elsewhere.

## themes
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  viridis::scale_color_viridis(discrete = TRUE) +
  theme_bw() +
  theme(legend.position = 'bottom')
  
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  viridis::scale_color_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = 'bottom')
```

* 1st query
  * **`theme()`**
   * `legend.position = 'bottom'`: move the legend from right (default) to bottom, to save place.
     * `legend.position = "none"`: remove the legend -- helpful when multiple plots use the same color scheme or when the legend is obnoxious for some other reason.
  * by default is `theme_grey`, but w `theme_bw()` we can change the grey background into white & change the white lines into black.
  * The ordering of `theme_bw()` and `theme()` matters – theme() changes a particular element of the plot’s current “theme”. If you call theme to change the some element and then theme_bw(), the changes introduced by theme() are overwritten by theme_bw().
* 2nd query
  * `theme_minimal()`: is theme_bw but w/o the box around the outside
* `theme_classic()`: no inside lines, and have outside lines on the bottom and the left of the plot.

## setting options
```{r}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

* use these in the 1st code trunk for every rmd to format all of the plots in a rmd. you can adjust if you want to.

## data argument
```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax)) +
  geom_point(aes(color = name)) +
  geom_smooth()
```

* where an aesthetic gets mapped to a variable matters
  * here, colored by name only applied to the scatterplot but not the smooth lines. if it locates into ggplot(), then this aes will define the whole bg, so both scatterplot and smooth lines would be applied to it.

## overlay data summaries on a plot of the complete data 
```{r}
nyc_weather_df =
  weather_df |> 
  filter(name == "CentralPark_NY")

hawaii_weather_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(nyc_weather_df, aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_line(data = hawaii_weather_df)
```
* we can have 1 df in 1 geometry and a diff df in a diff geometry
  * here, i add in hawaii ds as a completely separate ds sitting on top of the ggplot that i make from the beginning.
  
## patchwork

We can create the “same plot” for several levels of a categorical variable. But what to do if i need to make plots that are fundamentally diff in diff panels that i have? eg. juxtapose a scatterplot and a boxplot, or show scatterplots illustrating relationships between different variables.

* step 0: loading the patchwork package at the beginning
* step 1: create each of the panels you want separately
* step 2: combine panels using tools in the patchwork package:
```{r, message=FALSE}
library(patchwork)
```

```{r}
tmax_tmin_p = 
  weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name)) + 
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_dens_p = 
  weather_df |> 
  filter(prcp > 25) |> 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) + 
  theme(legend.position = "none")

tmax_date_p = 
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) + 
  theme(legend.position = "bottom")

tmax_tmin_p + prcp_dens_p

tmax_tmin_p + prcp_dens_p / tmax_date_p

(tmax_tmin_p + prcp_dens_p) / tmax_date_p
```
* **`tmax_tmin_p + prcp_dens_p`**: put these 2 plots together.
* **`tmax_tmin_p + prcp_dens_p / tmax_date_p`**: `put tmax_date_p` below `prcp_dens_p`, and then put them and `tmax_tmin_p` together.
* **`(tmax_tmin_p + prcp_dens_p) / tmax_date_p`**: put `tmax_tmin_p` and `prcp_dens_p` together, and then put `tmax_date_p` below them.

## data manipultaion

Factor variables are as categorical var that have an **order**.
```{r}
weather_df |> 
  mutate(
    name = fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))
  ) |> 
  ggplot(aes(x = name, y = tmax)) +
  geom_boxplot()

weather_df |> 
  mutate(
    name = fct_reorder(name, tmax)
  ) |> 
  ggplot(aes(x = name, y = tmax, color = name)) +
  geom_violin()
```

* **`fct_relevel()`**: r will regard character var as factor var and order them alphabetically (though it still shows in chr type). so we use `fct_relevel()` when we want to relevel the factor level in the way we like. but we need to do the typing manually.
* **`fct_reorder()`**: If we want to put name in the order according to some variables in our ds (eg. here i put name in order according to tmax var. r will compute the mean tmax for each group, and make the 1st factor the one w lowest mean tmax). if we have lots of factors, use `fct_reorder()` is better as we dont need to type all factors bu hand. 
  * we can do this before creating plots. no need to create a new factor var unless necessary.
  
## complicate FAS plot

### example 1
```{r}
litters_df =
  read_csv("data/FAS_litters.csv") |> 
  janitor::clean_names() |> 
  separate(group, into = c('dose', 'day_of_tx'), sep = 3)

pup_data = 
  read_csv("./data/FAS_pups.csv") |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"))

fas_df =
  left_join(pup_data, litters_df, by = 'litter_number')

fas_df |> 
  select(pd_ears:pd_walk, day_of_tx, dose) |> 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = 'outcome',
    values_to = 'pn_day'
  ) |> 
  drop_na() |> 
  mutate(
    outcome = fct_reorder(outcome, pn_day)
  ) |> 
  ggplot(aes(x = dose, y = pn_day)) +
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome)

```
* `separate(group, into = c('dose', 'day_of_tx'), sep = 3)`: separate group var into 'dose' and 'day_of_tx'. `sep = 3` means separate at the 3rd character.
* we can use either `pd_ears:pd_walk` or `starts_with('pd')` as these 4 cols we want all have same prefix of 'pd'

### example 2

A more difficult situation relates to data tidiness. Suppose I wanted to create a three-panel plot showing densities for tmax and tmin within each location. More concretely, I want to be able to facet panels across the name variable, and create separate densities for tmax and tmin in each panel. Unfortunately, weather_df isn’t organized in a way that makes this easy.

One solution would recognize that tmax and tmin are separate observation types of a shared temperature variable. With this understanding, it’s possible to tidy the weather_df (using `pivot_longer`) and make the plot directly:
```{r}
weather_df |>
  select(name, tmax, tmin) |> 
  pivot_longer(
    tmax:tmin,
    names_to = "observation", 
    values_to = "temp") |> 
  ggplot(aes(x = temp, fill = observation)) +
  geom_density(alpha = .5) + 
  facet_grid(~name) + 
  viridis::scale_fill_viridis(discrete = TRUE)
```

### example 3

Our emphasis on data tidiness in previous examples is helpful in visualization. The code below imports and tidies the PULSE data, and creates a plot showing BDI score across visits. Some steps that are helpful in retrospect are using `pivot_longer` to organize the BDI score and visit time variables, and *organizing the visit time variable into a factor with an informative ordering*.
```{r}
pulse_data = 
  haven::read_sas("data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  select(id, visit, everything()) |>
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) |>
  arrange(id, visit)

ggplot(pulse_data, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```

* can also use `visit = fct_relevel(visit, c('00m', '01m', '06m', '12m'))` here instead of `visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))`
