---
title: "Visualization part 2"
author: "Shihui Peng"
date: "2023-10-12"
output: github_document
---
```{r, message=FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

# Get the data for plotting 
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

## labeling w `labs()`
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  )
```
* **`labs()`**: label, and can label lots of aesthetics. inside the `labs()`:
  * `x =` & `y =`: label the x and y axis.
  * `color =`: we have a color aeshetics mapped onto the name var, so the name var show up as the legend of the color bar. now, i want to label color as 'location'.
  * `title=` and `caption=` to define title and caption
  
## scales

we have basically 1 to 1 mapping of whatever on the aes to whatever var that are put in. but we might want to do some tramsformation in some association. we can start from adding diff scales to it to modify the default mapping bt a var and the aes
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  scale_x_continuous(
    breaks = c(-15, 0, 15),
    labels= c('-15C', '0C', '15C')
  ) +
  scale_y_continuous(
    position = 'right',
    trans = 'sqrt',
    limit = c(20, 30)
  )
```

* we can control over the location and specification of tick marks on the X or Y axis w **`scale_x_* and scale_y_*`** where * depends on the type of variable mapped to the x and y aesthetics (i.e. continuous vs discrete).
  * **`scale_x_continuous()`**: 
    * `breaks = c(-15, 0, 15)`: use this to control where things fall on the x-axis
    * `labels =` is to modify the labels for each point falls o the x-axis.
  * **`scale_y_continuous()`**:
    * `position = 'right'`: move the info of y-axis from left (default) to right.
    * `trans = 'sqrt'`: do a square root transformation. we can also do this w `mutate` before ggplot, and this can be embedded to further down below.
    * `limit = c(20, 30)`: limit the y-axis value wi 20~30. but `filter()` might be better, bc w limit=, x-axis here is not updated and is still the corresponding range with the previous y range.
    
## colors
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  labs(
    x = "Min daily temp (degree C)",
    y = "Max daily temp (degree C)",
    color = "location",
    title = "temperature plot",
    caption = "Max vx min daily temp in 3 location; data from rnoaa"
  ) +
  viridis::scale_color_viridis(discrete = TRUE)
  
```
* we can use `scale_color_hue(h = c(150, 300))` to create our own color scheme, but it usually does not go well. So use **viridis** package is better.
* We used `discrete = TRUE` because the color aesthetic is mapped to a discrete variable. In other cases (for example, when color mapped to prcp) you can omit this argument to get a continuous color gradient. The `viridis::scale_fill_viridis()` function is appropriate for the fill aesthetic used in histograms, density plots, and elsewhere.

## themes
```{r}

```

